---
title: "HDPSA Milestone 1: Data Exploration and Data Dictionary"
subtitle: "Business Intelligence 371"
author: "Member 1"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(readxl)
library(tidyverse)
library(skimr)
library(janitor)
library(knitr)
```

# Load all datasets

Put the 10 xlsx files in `data/raw/`. This reads every file into a named list,
so you never hard code file names.

```{r load}
data_dir <- "data/raw"

files <- list.files(data_dir, pattern = "\\.xlsx$", full.names = TRUE)

datasets <- files |>
  set_names(~ .x |> basename() |> str_remove("_national_zaf.xlsx")) |>
  map(read_excel)

tibble(
  dataset = names(datasets),
  rows    = map_int(datasets, nrow),
  columns = map_int(datasets, ncol)
) |>
  kable(caption = "Datasets loaded")
```

# Structure and summary per dataset

```{r structure, results='asis'}
for (nm in names(datasets)) {
  cat("\n\n## ", nm, "\n\n")
  df <- datasets[[nm]]
  cat("Dimensions:", nrow(df), "rows x", ncol(df), "columns\n\n")
  print(kable(skim(df)))
}
```

# Data quality checks

The teaching data has planted problems. This section finds them automatically
so nothing is missed. The columns below are rates or times, not 0 to 100
percentages, so they are excluded from the "over 100" check.

```{r quality-config}
not_percentage <- c(
  "RecordID", "SurveyYear",
  "Under5MortalityRate", "InfantMortalityRate",
  "MaternalMortalityRate", "CollectionTime"
)
```

## Missing values

```{r missing}
missing_report <- imap_dfr(datasets, function(df, name) {
  tibble(
    dataset     = name,
    variable    = names(df),
    n_missing   = map_int(df, ~ sum(is.na(.x))),
    pct_missing = round(map_dbl(df, ~ mean(is.na(.x)) * 100), 1)
  )
}) |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing))

kable(missing_report, caption = "Columns with missing values")
```

## Duplicate rows

```{r duplicates}
imap_dfr(datasets, ~ tibble(dataset = .y, duplicate_rows = sum(duplicated(.x)))) |>
  kable(caption = "Duplicate row count per dataset")
```

## Inconsistent categories

Flags text columns where the same value appears in different letter cases,
for example urban vs Urban.

```{r categories, results='asis'}
for (nm in names(datasets)) {
  df <- datasets[[nm]]
  text_cols <- names(df)[map_lgl(df, is.character)]
  text_cols <- setdiff(text_cols, "Province")  # Province is consistent
  for (col in text_cols) {
    vals <- df[[col]]
    if (n_distinct(vals) != n_distinct(str_to_lower(vals))) {
      cat("\n\n**", nm, "->", col, "has case inconsistencies:**\n\n")
      print(kable(as.data.frame(table(vals))))
    }
  }
}
```

## Impossible values

Negative values in any measure, and percentages above 100.

```{r invalid}
invalid_report <- imap_dfr(datasets, function(df, name) {
  num_cols <- names(df)[map_lgl(df, is.numeric)]
  map_dfr(num_cols, function(col) {
    x <- df[[col]]
    n_negative <- sum(x < 0, na.rm = TRUE)
    n_over_100 <- if (col %in% not_percentage) 0L else sum(x > 100, na.rm = TRUE)
    tibble(dataset = name, variable = col, n_negative, n_over_100)
  })
}) |>
  filter(n_negative > 0 | n_over_100 > 0)

kable(invalid_report, caption = "Impossible values found")
```

# Temporal analysis: 1998 vs 2016

Mean of every numeric indicator by survey year, per dataset.

```{r temporal, results='asis'}
for (nm in names(datasets)) {
  df <- datasets[[nm]]
  summary_tbl <- df |>
    group_by(SurveyYear) |>
    summarise(across(where(is.numeric) & !RecordID, ~ round(mean(.x, na.rm = TRUE), 2)),
              .groups = "drop")
  cat("\n\n## ", nm, "\n\n")
  print(kable(summary_tbl))
}
```

# Correlations

Correlation between numeric indicators within each dataset (ID and year removed).

```{r correlation, results='asis'}
for (nm in names(datasets)) {
  df <- datasets[[nm]] |>
    select(where(is.numeric), -any_of(c("RecordID", "SurveyYear"))) |>
    drop_na()
  if (ncol(df) >= 2) {
    cat("\n\n## ", nm, "\n\n")
    print(kable(round(cor(df), 2)))
  }
}
```

# Distributions

Histograms of every numeric indicator, one facet grid per dataset.

```{r distributions, results='asis', fig.width=9, fig.height=5}
for (nm in names(datasets)) {
  df <- datasets[[nm]] |>
    select(where(is.numeric), -any_of(c("RecordID", "SurveyYear"))) |>
    pivot_longer(everything(), names_to = "variable", values_to = "value")
  cat("\n\n## ", nm, "\n\n")
  p <- ggplot(df, aes(value)) +
    geom_histogram(fill = "steelblue", colour = "white", bins = 30) +
    facet_wrap(~ variable, scales = "free") +
    theme_minimal() +
    labs(title = paste("Distributions:", nm), x = NULL, y = "Count")
  print(p)
}
```

# Data dictionary

Auto builds most columns. Fill description, unit and valid_range by hand using
the DHS indicator definitions, then export.

```{r dictionary}
build_dict <- function(df, source_name) {
  tibble(
    dataset     = source_name,
    variable    = names(df),
    type        = map_chr(df, ~ class(.x)[1]),
    n_missing   = map_int(df, ~ sum(is.na(.x))),
    pct_missing = round(map_dbl(df, ~ mean(is.na(.x)) * 100), 1),
    n_unique    = map_int(df, ~ n_distinct(.x)),
    example     = map_chr(df, ~ paste(head(na.omit(unique(.x)), 3), collapse = ", ")),
    description = "",
    unit        = "",
    valid_range = ""
  )
}

dictionary <- imap_dfr(datasets, build_dict)
dir.create("outputs", showWarnings = FALSE)
write_csv(dictionary, "outputs/data_dictionary.csv")
kable(head(dictionary, 20), caption = "Data dictionary (first 20 rows)")
```

# Quality summary for Member 2

```{r handoff}
quality_summary <- imap_dfr(datasets, function(df, name) {
  tibble(
    dataset        = name,
    rows           = nrow(df),
    columns        = ncol(df),
    total_missing  = sum(is.na(df)),
    duplicate_rows = sum(duplicated(df))
  )
})
write_csv(quality_summary, "outputs/quality_summary.csv")
kable(quality_summary, caption = "Quality summary across all datasets")
```

# Session info

```{r session}
sessionInfo()
```
